---
title: "Template"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
# install.packages("tidyverse")
# install.packages("yaml")
# install.packages("rmarkdown")
# install.packages("dplyr")
# install.packages("ggplot2")

```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
library(dplyr)
```

# Title Page

Naut Leuverink, Lucas Wolffenbuttel, Mike Ottervanger, Otto Ruyten, Nicolaas van Nispen en Robin Koldenhof

4.3

Fatima Amankour

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Include the following:

- Why is this relevant?

- Nitrogen emissions represent a critical public health crisis in the Netherlands. Elevated atmospheric concentrations lead to the production of particulate manner and ozone, which are closely associated with serious health problems such as worsening asthma, long-term cardiovascular problems, and diminished lung function.

- This report measures the correlation between public health concerns and regional nitrogen levels over almost 20 years, from 2005 to 2023. By analyzing these temporal and spatial patterns, we want to pinpoint pollution hotspots and offer a data-driven framework for focused reduction initiatives, like industrial restrictions or localized agriculture changes, to successfully lessen the burden on public health.

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}
PolutionInGeneral<- read.csv("data/Polution in general.csv")
BA_BronnenVanStikstof <- read.csv("data/Bronnen van stikstof.csv")
AA_BronnenVanFijnstof <- read.csv("data/Bronnen van Fijnstof.csv")
JaargemiddeldeFijnstof <- read.csv("data/Jaargemiddelde PM10 per Locatie 2015-2025.csv", sep = ";")

saveRDS(PolutionInGeneral, "data/PolutionInGeneral.rds")
PolutionInGeneral < readRDS("data/PolutionInGeneral.rds")
saveRDS(BA_BronnenVanStikstof, "data/BA_BronnenVanStikstof.rds")
BA_BronnenVanStikstof <- readRDS("data/BA_BronnenVanStikstof.rds")
saveRDS(AA_BronnenVanFijnstof, "data/AA_BronnenVanFijnstof.rds")
AA_BronnenVanFijnstof <- readRDS("data/AA_BronnenVanFijnstof.rds")
saveRDS(JaargemiddeldeFijnstof, "data/JaargemiddeldeFijnstof.rds")
JaargemiddeldeFijnstof <- readRDS("data/JaargemiddeldeFijnstof.rds")
```

## 2.2 Provide a short summary of the dataset(s)

```{r}
#head(PolutionInGeneral)
```

This dataset contains 11 variables, each representing a different year. For each year, data is provided on the levels of nitrogen, NMVOCs (non-methane volatile organic compounds), sulfur oxides, and PM2.5 particulate matter observed in the air. By analyzing this data, we can assess the severity of air pollution and identify trends in air quality over time.

```{r}
#head(JaargemiddeldeFijnstof)
```

These datasets show particulate matter (fine dust) levels measured at various monitoring stations across the Netherlands over the years 2005 till 2025. This provides a clear insight into the periods and regions where particulate matter causes the greatest impact and pollution.

```{r}
#head(BA_BronnenVanStikstof)
```

This dataset shows the sources of nitrogen measured over a time period from 2005 till 2024

```{r}
#head(AA_BronnenVanFijnstof)
```

This dataset shows the sources of fine dust measured over a time period from 2005 till 2024

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Most of the datasets include a duration of time most often 2005 till 2024, and a variable that may be a specific measuring station or a source of polution.

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}

#Laatste rij was .cbs, die hebben we verwijderd
AA_BronnenVanFijnstof<- AA_BronnenVanFijnstof [-nrow(AA_BronnenVanFijnstof),]

AA_BronnenVanFijnstof <- AA_BronnenVanFijnstof %>%
  mutate(across(where(is.character),~ as.numeric(gsub(",",".", .x))))

#Nieuwe variabele gemaakt 
AA_BronnenVanFijnstof$Totaal <- AA_BronnenVanFijnstof$Landbouw..bosbouw.en.visserij..mln.kg.+ AA_BronnenVanFijnstof$Particuliere.huishoudens..mln.kg.+ AA_BronnenVanFijnstof$Energiesector..mln.kg.+ AA_BronnenVanFijnstof$Industrie.en.afval..mln.kg.+ AA_BronnenVanFijnstof$Vervoer..mln.kg.+ AA_BronnenVanFijnstof$Diensten..water.en.bouw..mln.kg.

```

```{r}
BA_BronnenVanStikstof<- BA_BronnenVanStikstof [-nrow(BA_BronnenVanStikstof),]

BA_BronnenVanStikstof <- BA_BronnenVanStikstof %>%
  mutate(across(where(is.character),~ as.numeric(gsub(",",".", .x))))


BA_BronnenVanStikstof$Totaal <- BA_BronnenVanStikstof$Landbouw..bosbouw.en.visserij..mln.kg.+ BA_BronnenVanStikstof$Particuliere.huishoudens..mln.kg.+ BA_BronnenVanStikstof$Energiesector..mln.kg.+ BA_BronnenVanStikstof$Industrie.en.afval..mln.kg.+ BA_BronnenVanStikstof$Vervoer..mln.kg.+ BA_BronnenVanStikstof$Diensten..water.en.bouw..mln.kg.
```

2 datasets naast elkaar gezet, en dat per jaar vergeleken, dataset fijnstof gaat tot 2024 en dataset stikstof gaat tot 2023, daar moest ik nog een aanpassing in maken.

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1
### Creates total variables

```{r}
# De bestaande kolom 'Totaal' wordt gebruikt.
# De totalen van fijnstof en stikstof worden gecombineerd op basis van het jaar.
# Omdat stikstof tot en met 2023 loopt, loopt de gecombineerde tabel ook tot en met 2023.

AB_Fijnstof_Totaal <- data.frame(
  Jaar = AA_BronnenVanFijnstof[, 1],
  Totaal_fijnstof = AA_BronnenVanFijnstof$Totaal
)

BB_Stikstof_Totaal <- data.frame(
  Jaar = BA_BronnenVanStikstof[, 1],
  Totaal_stikstof = BA_BronnenVanStikstof$Totaal
)

totale_data <- merge(
  AB_Fijnstof_Totaal,
  BB_Stikstof_Totaal,
  by = "Jaar"
)

totale_data
```

## 3.3 Visualize temporal variation

### Simple graph about the general emission of particulate matter (fijnstof)

```{r}
library(ggplot2)

ggplot(AA_BronnenVanFijnstof, aes(x = X., y = Totaal)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Total particulate matter emissions over the years",
    x = "Years",
    y = "Particulate matter (mln kg)"
  ) +
  theme_minimal()
```

### Simple graph about the general emission of nitrogen

```{r}
library(ggplot2)

ggplot(BA_BronnenVanStikstof, aes(x = X., y = Totaal)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Total nitrogen emissions over the years",
    x = "Years",
    y = "Nitrogen (mln kg)"
  ) +
  theme_minimal()
```

### Complicated and extensive graph of the nitrogen emissions

```{r}
# 1. Laad de benodigde bibliotheken
library(tidyverse)

# 2. Bereid de dataset voor: hernoem de eerste kolom naar "X."
names(BA_BronnenVanStikstof)[1] <- "X."

# 3. Zet de tabel om naar het lange formaat dat jouw ggplot verwacht
BA_BronnenVanStikstof_long <- BA_BronnenVanStikstof %>%
  pivot_longer(
    cols = -X., 
    names_to = "Bron", 
    values_to = "Uitstoot"
  ) %>%
  # BELANGRIJK: Filter de "Totaal"-kolom eruit zodat deze de grafiek niet vertekent
  filter(Bron != "Totaal")


ggplot(BA_BronnenVanStikstof_long,
       aes(x = X.,                                    # <--- Werkt nu met de jaren uit je CSV
           y = Uitstoot, 
           fill = Bron)) +
  geom_area(color = "white", linewidth = 0.2) +       # <--- Creëert de witte scheidingslijntjes
  scale_fill_brewer(palette = "Set3") +
  labs(                                               # <--- Namen op de assen en titels
    title = "Nitrogen emissions per source",    
    x = "Years",
    y = "Emissions",
    fill = "Sources"
  ) +
  theme_bw() +                                        # <--- Grijze rasterlijnen inschakelen
  theme(
    panel.grid.major = element_line(color = "grey80"),
    panel.grid.minor = element_line(color = "grey90"),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

```{r}
# 1. Laad de benodigde bibliotheken
library(tidyverse)

# 2. Bereid de dataset voor: hernoem de eerste kolom naar "X."
names(AA_BronnenVanFijnstof)[1] <- "X."

# 3. Zet de tabel om naar het lange formaat dat jouw ggplot verwacht
AA_BronnenVanFijnstof_long <- AA_BronnenVanFijnstof %>%
  pivot_longer(
    cols = -X., 
    names_to = "Bron", 
    values_to = "Uitstoot"
  ) %>%
  # BELANGRIJK: Filter de "Totaal"-kolom eruit zodat deze de grafiek niet vertekent
  filter(Bron != "Totaal")


ggplot(AA_BronnenVanFijnstof_long,
       aes(x = X.,                                    # <--- Werkt nu met de jaren uit je CSV
           y = Uitstoot, 
           fill = Bron)) +
  geom_area(color = "white", linewidth = 0.2) +       # <--- Creëert de witte scheidingslijntjes
  scale_fill_brewer(palette = "Set3") +
  labs(                                               # <--- Namen op de assen en titels
    title = "Nitrogen emissions per source",    
    x = "Years",
    y = "Emissions",
    fill = "Sources"
  ) +
  theme_bw() +                                        # <--- Grijze rasterlijnen inschakelen
  theme(
    panel.grid.major = element_line(color = "grey80"),
    panel.grid.minor = element_line(color = "grey90"),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

## 3.4 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document. For the datasets: AA_BronnenVanFijnstof, BA_BronnenVanStikstof and PolutionIngeneral are from CBS:

Centraal Bureau voor de Statistiek. (2025, 2 september). Uitstoot luchtverontreinigende stoffen lager dan norm voor 2030. <https://www.cbs.nl/nl-nl/nieuws/2025/36/uitstoot-luchtverontreinigende-stoffen-lager-dan-norm-voor-2030> Dataset JaargemiddeldeFijnstof Planbureau voor de Leefomgeving. (z.d.). Atlas van de Regio. <https://themasites.pbl.nl/atlas-regio/>
