---
title: "Template"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
# install.packages("tidyverse")
#install.packages("yaml")
#install.packages("rmarkdown")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
```

# Title Page

Naut Leuverink, Lucas Wolffenbuttel, Mike Ottervanger, Otto Ruyten, Nicolaas en Robin Koldenhof

4.3

Fatima Amankour

# Part 1 - Identify a Social Problem


Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Include the following:

-   Why is this relevant?

-   ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}
PolutionInGeneral<- read.csv("data/Polution in general.csv")
BronnenVanStikstof <- read.csv("data/Bronnen van stikstof.csv")
BronnenvanFijnstof <- read.csv("data/Bronnen van Fijnstof.csv")
JaargemiddeldeFijnstof <- read.csv("data/Jaargemiddelde PM10 per Locatie 2015-2025.csv", sep = ";")

saveRDS(PolutionInGeneral, "data/PolutionInGeneral.rds")
PolutionInGeneral < readRDS("data/PolutionInGeneral.rds")
saveRDS(BronnenVanStikstof, "data/BronnenVanStikstof.rds")
BronnenVanStikstof <- readRDS("data/BronnenVanStikstof.rds")
saveRDS(BronnenvanFijnstof, "data/BronnenvanFijnstof.rds")
BronnenvanFijnstof <- readRDS("data/BronnenVanFijnstof.rds")
saveRDS(JaargemiddeldeFijnstof, "data/JaargemiddeldeFijnstof.rds")
JaargemiddeldeFijnstof <- readRDS("data/JaargemiddeldeFijnstof")
```

## 2.2 Provide a short summary of the dataset(s)

```{r}
head(PolutionIngeneral)
```

This dataset contains 11 variables, each representing a different year. For each year, data is provided on the levels of nitrogen, NMVOCs (non-methane volatile organic compounds), sulfur oxides, and PM2.5 particulate matter observed in the air. By analyzing this data, we can assess the severity of air pollution and identify trends in air quality over time.

```{r}
head(JaargemiddeldeFijnstof)
```

These datasets show particulate matter (fine dust) levels measured at various monitoring stations across the Netherlands over the years 2005 till 2025. This provides a clear insight into the periods and regions where particulate matter causes the greatest impact and pollution.

```{r}
head(BronnenVanStikstof)
```

This dataset shows the sources of nitrogen measured over a time period from 2005 till 2024

```{r}
head(BronnenvanFijnstof)
```

This dataset shows the sources of fine dust measured over a time period from 2005 till 2024


``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Most of the datasets include a duration of time most often 2005 till 2024, and a variable that may be a specific measuring station or a source of polution. 


# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}
mean(dataset$percollege)
```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc...

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}

```

Variable 2

```{r gen_var2}

```

## 3.3 Visualize temporal variation

```{r}

```

## 3.4 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
